---
title: "Final Project"
author: "Steve Dunn"
date: "July 17, 2016"
output: html_document
---
Prof. Andy Catlin
IS 643 - Special Topics: Recommender Systems
Department of Data Analytics, City University of New York
```{r, include=FALSE}
# load requried libraries and data
library(ggplot2)
library(plyr)
library(RColorBrewer)
library(grid)
library(reshape2)
library(caret)
library("data.table")
library(pander)
library("recommenderlab")
library("lsa")
library(Metrics)
data("MovieLense")
```

For my Final project I will build a recommendation system that will look at recommendation from Item Based Collaborative Filtering, User Based Collaborative Filtering and recommendations by genre with Hierarchal clustering. I will create models to get Movies from similar Genres based on the distance of the clusters.Hierarchal Clustering is an unsupervised learning method where the goal is to segment data into similar groups.The dataset is taken from the MovieLens dataset avaliable in the Recommenderlab package and from their website http://grouplens.org/datasets/movielens.

#Data Exploration

```{r}
#Visualization of the data
str(MovieLense)
class(MovieLense)
head(MovieLense)
dim(MovieLense)
summary(MovieLense)

## visualize part of the matrix
image(MovieLense[1:100,1:100])

## number of ratings per user
hist(rowCounts(MovieLense))

## number of ratings per movie
hist(colCounts(MovieLense))

## mean rating (averaged over users)
mean(rowMeans(MovieLense))
```

**calculate similarity between user ratings and item ratings**

the red areas denote the similarity of the ratings,the diagonals are all red because it is measured against it self.There seems to be more similarity between user ratings than there is between items.

```{r}
#calculate Similarity matrix
similarity_users <- similarity(MovieLense[1:4,], method = "cosine", which = "users")
as.matrix( similarity_users)
image(as.matrix(similarity_users), main = "User similarity")

  similarity_items <- similarity(MovieLense[,1:4], method = "cosine", which = "items")
as.matrix(similarity_items)
image(as.matrix(similarity_items), main = "Movies similarity")
```

splitting the data set into training and test where i will use the training set to rate the data and use the test data set to predict the movies by IBCF and UBCF

```{r}
#=============================================================================================================
#Splitting the dataset into Train and test
#=============================================================================================================
#Subsetting the dataset
ratings_movies <- MovieLense[ rowCounts( MovieLense) > 50, colCounts( MovieLense) > 100]
ratings_norm<-(ratings_movies)

which_train<-sample(x=c(TRUE,FALSE),size = nrow(ratings_norm),replace = TRUE,prob = c(.8,.2))
rec_data_train<-ratings_movies[which_train,]
rec_data_test<-ratings_movies[!which_train,]

```

***Item Based Collaborative Filtering***

```{r}

#Prediction for Item Based Collaborative Filtering

#build recommender Model
rec_modelI<-Recommender(data=rec_data_train,method='IBCF')

#predict 5 movies for a specific users
n_rec<-5
recom2<-predict(rec_modelI,newdata=rec_data_test[1],n=n_rec)
#convert recommenderlab object to readable list
recom_list <- as(recom2, "list") 
recom_list

#predict 5 movies for users
rec_predictI<-predict(rec_modelI,newdata=rec_data_test,n=n_rec)
rec_predictI
rec_matrix<-sapply (rec_predictI@items,
function(x)
  {
  colnames(rec_data_test)[x]
  }
)
rec_matrix[,1:5]
```

***User Based Collaborative Filtering***

```{r}
#==========================================================================================
#Prediction for User Based Collaborative Filtering
#==========================================================================================
#build recommender Model
rec_modelU<-Recommender(data=rec_data_train,method='UBCF')
model_details <- getModel(rec_modelU)

#predict 5 movies for specific users
n_rec<-10

#Obtain 10 recommendations for 1st user in dataset
recom <- predict(object=rec_modelU, newdata=rec_data_test[1], n=n_rec) 

#output ot a list
recom_list <- as(recom, "list") 
recom_list

#predict  movies for users
rec_predictU<-predict(rec_modelU,newdata=rec_data_test,n=n_rec)
rec_predictU

rec_matrixU<-sapply(rec_predictU@items,
                    function(x){
                      colnames(rec_data_test)[x]
                      })
rec_matrixU[,1:5]
```

Using the built in functions from RecommenderLab to evaluate the models,Using K fold validation it can be seen that UBCF performs better than IBCF.

```{r}
#Model Evaluation
#A 5-fold cross validation
n_fold=5
evaluation_scheme <- evaluationScheme(ratings_movies, method="cross-validation", n_fold, given=3, goodRating=5) 
evaluation_results <- evaluate(evaluation_scheme, method="UBCF")
eval_results <- getConfusionMatrix(evaluation_results)[[1]]
eval_results #The evaluation results of the top recommendation

algorithms <- list(
  IBCF = list(name = "IBCF", param = NULL),
  UBCF = list(name = "UBCF", param = NULL),
  SVD = list(name = "SVD", param = NULL)
)

evlist <- evaluate(evaluation_scheme, algorithms)

plot(evlist, legend="topleft")

plot(evlist,"prec", legend="bottomright")

plot(evaluation_results,"prec/rec", annotate = TRUE, main = "Precision-recall")
```

###Hierarchial Cluster by Genres

The data has to be re-organize so that it allows for the movies to be displayed by specific genres, this is much easier compared to selecting a movie from a large list of all available movies

```{r, include=FALSE}
#Recommendation by Genres

#content Based Filtering
movies = read.table("~/SPS/IS643/project 4/movies.txt", header=FALSE, sep="|",quote="\"")

colnames(movies) = c("ID", "Title", "ReleaseDate", 
                     "VideoReleaseDate", "IMDB", "Unknown",
                     "Action", "Adventure", "Animation",
                     "Childrens", "Comedy", "Crime", 
                     "Documentary", "Drama", "Fantasy",
                     "FilmNoir", "Horror", "Musical", 
                     "Mystery", "Romance", 
                     "SciFi", "Thriller", "War", "Western")

#remove unwnated columns
movies$ID = NULL
movies$ReleaseDate = NULL
movies$VideoReleaseDate = NULL
movies$IMDB = NULL

#Remove duplicated MOvies
movies = unique(movies)

```

Take a subset from the data and calculate the distance between genres,using the Euclidean method which measures the distance between each pair of points within each cluster .After calculating the distance I used a dendogram to display the data with average,complete and ward methods.The distance matrix shows the distance between clusters e.g. the distance between cluster 3 and 20 is 1.73 so this indicates that there are possible good similarity of movies between these two clusters.The larger the distance the more likely that genres are disimilar. 

_Initially each genre is trated as a single cluster and then the algorithm tries to find the closest in distance until all the genres form one large cluster_  

```{r}
set.seed(1306)
sammovies <- movies[sample(2:20, 11),] # create sample from the data set

dstncs<- dist(sammovies[2:20], method="euclidean") #The distance between two points 

print(dstncs,digits=3)#euclidean distance between each variable

cluster<- hclust(dstncs)#dendogram 

cluster<- hclust(dstncs, method="average")#dendogram with average 

#The vertical lines represents the distance between clusters
plot(cluster, cex = 0.6)
plot(cluster, labels=sammovies$Title,cex = 0.6)
```

**_Displaying cluster with Ward.D2  method taking a look at the movies Toy Story and Psycho to see which other movies are similar to them and the genres that they belong to_**

_The Euclidean distance between points p and q is the length of the line segment connecting them_

```{r}
clustMovies = hclust(dstncs, method = "ward.D2")
cm=hclust(dstncs, method = "ward.D2")

plot(cm, cex = 0.6) 

plot(clustMovies,cex = 0.6)

plot(clustMovies, labels=sammovies$Title,cex = 0.6)

#cuts the dendogram into 10 groups
clustGrps1=cutree(clustMovies, k=10)

# Calculate distances between genre features
distances = dist(movies[2:20], method = "euclidean")

clustMovies = hclust(distances, method = "ward.D2")

#plot(clustMovies, hang=-1,cex = 0.6)

#Label each movie in the clusters 
clustGrps = cutree(clustMovies, k=10)
```

```{r}
#compute the average value and percentage of movies in each genre and for each cluster 
tapply(movies$Comedy, clustGrps, mean)
tapply(movies$Horror, clustGrps, mean)
```

```{r}
subset(movies, Title=="Toy Story (1995)")# get the cluster that Toy story and Psycho belongs to 
cluster2 = subset(movies, clustGrps==1)
cluster2$Title[1:10]

subset(movies, Title=="Psycho (1960)")
cluster3 = subset(movies, clustGrps==3)
cluster3$Title[1:10]
```

Recommending movies with similar genres allows for a possible wider choice in selection as opposed to UBCF where the recommendation is based on other users rating.Hierarchialclustering usually starts with each variable in its own cluster and then combines the next closest cluster by euclidean method until their is one large cluster.

***_weighting of variable within the cluster_***

```{r}
#calculate the cluster means
binsp = split(movies[2:20], clustGrps)
lapply(binsp, colMeans)
```
