---
title: "SD_Assignment 2"
author: "Steve Dunn"
date: "June 26, 2016"
output: html_document
---
Prof. Andy Catlin
IS 643 - Special Topics: Recommender Systems
Department of Data Analytics, City University of New York
```{r setup, include=FALSE}
library("recommenderlab")
library("ggplot2")
library("lsa")
library(caret)
data("MovieLense")
table_Movie <- read.csv("~/SPS/IS643/Movie.csv")
table_Users <- read.csv("~/SPS/IS643/Users.csv")
(echo = FALSE)
```
#======================================================================================
###Using R package to build the similarity and  Collaborative Filtering recommender System code and data set Mostly taken from example in Building a Recommendation System with R
#======================================================================================

While the models use for recommendations do as is expected i.e. make recommendation the accuracy and precision of those recommendation have to be evaluated to see if they make sense either at a user -user or item-item filtering. The accuracy is measured with RMSE, MSE and MAE. Since the measure as calculated differently and apply different weightings I would choose between the RMSE or MAE.I also found that using k-fold cross validation is faster than splitting the data set manually.
```{r,echo=FALSE}
## visualize part of the matrix
image(MovieLense[1:100,1:100])

## number of ratings per user
hist(rowCounts(MovieLense))

## number of ratings per movie
hist(colCounts(MovieLense))

## mean rating (averaged over users)
mean(rowMeans(MovieLense))
```

```{r,echo=FALSE}
sim_user<-similarity(MovieLense[1:10,], method='cosine',which='users')
as.matrix(sim_user)
image(as.matrix(sim_user),main="Similarity of Users")
head(as.matrix(sim_user)[,1:10])

sim_item<-similarity(MovieLense[,1:10], method = 'cosine',which='items')
as.matrix(sim_item)
image(as.matrix(sim_item),main="Similarity of Items")
head(as.matrix(sim_item)[,1:5])
```
#Splitting the Data set into train and test
```{r}
#Subsetting the dataset
ratings_movies <- MovieLense[ rowCounts( MovieLense) > 50, colCounts( MovieLense) > 100]

which_train<-sample(x=c(TRUE,FALSE),size = nrow(ratings_movies),replace = TRUE,prob = c(.8,.2))
rec_data_train<-ratings_movies[which_train,]
rec_data_test<-ratings_movies[!which_train,]

#build recommender Model
rec_model<-Recommender(data=rec_data_train,method='IBCF')

#predict 10 movies for the users
n_rec<-10
rec_predict<-predict(rec_model,newdata=rec_data_test,n=n_rec)

rec_predict
rec_matrix<-sapply(rec_predict@items,
                   function(x){
                     colnames(rec_data_test)[x]
                   })
rec_matrix[,1:10]
```
#evaluating Recommendationusing k fold Cross Validation
```{r}

n_fold <- 10 
items_to_keep <- 15
rating_threshold <- 3
model_to_evaluate <- "IBCF" 
model_parameters <- NULL
```

```{r}
eval_sets <- evaluationScheme( data = ratings_movies,
                               method = "cross-validation"
                               , k = n_fold, given = items_to_keep
                               , goodRating = rating_threshold)

eval_prediction <- predict(object = rec_model,
                            newdata = getData( eval_sets, "known"),
                            n = 10, type = "ratings")

eval_accuracy <- calcPredictionAccuracy(
  x = eval_prediction,
  data = getData(eval_sets, "unknown"),
  byUser = FALSE,
  given = items_to_recommend)
head( eval_accuracy)

eval_recommender <- Recommender(data = getData(eval_sets, "train"),
                                method = model_to_evaluate,
                                parameter = model_parameters)

results <- evaluate(x = eval_sets,
                    method = model_to_evaluate,
                    n = seq(10, 100, 10))
head( getConfusionMatrix( results)[[ 1]])

```

```{r}
table_wide <- reshape( data =table_Users,#convert table to wide
                       direction = "wide",
                       idvar ="UserID",
                       timevar = "MovieID",
                       v.names = "Rating" )

###Item Based Collaborative Filtering
head(table_wide[, 1: 10])
W_movies<-table_wide
W_movies <- (table_wide[,!(names(table_wide) %in% c("UserID"))])
W_movies[is.na(W_movies)] = 0
head(W_movies[, 1:10])

# Function to calculate cosine Similarity
Sim_Cos <- function(a,b) 
{
  cos_ine <- sum(a*b) / (sqrt(sum(a*a)) * sqrt(sum(b*b)))
  return(cos_ine)
}

#Fill table with item rating
W_movies_sim  <- matrix(NA, nrow=ncol(W_movies)
                          ,ncol=ncol(W_movies)
                          ,dimnames=list(colnames(W_movies)
                          ,colnames(W_movies)))

for(i in 1:ncol(W_movies)) {
  # Loop through the columns 
  for(j in 1:ncol(W_movies)) {
    # Fill matrix with cosine similarities
    W_movies_sim[i,j] <- Sim_Cos(as.matrix(W_movies[i]),as.matrix(W_movies[j]))
  }
}
```

```{r}
image(as.matrix(W_movies_sim),main="Similarity of Items")

W_movies_sim <- as.data.frame(W_movies_sim)

W_movies_n <- matrix(NA, nrow=ncol(W_movies_sim),ncol=5,dimnames=list(colnames(W_movies_sim)))

for(i in 1:ncol(W_movies)) 
{
  W_movies_n[i,] <- (t(head(n=5,
              rownames(W_movies_sim
            [order(W_movies_sim[,i]
          ,decreasing=TRUE),][i]))))
}

#convert to rating Matrix
ratings_movies <- as(W_movies_sim, "realRatingMatrix")
ratings_movies

#Splitting the Data set into train and test
which_train<-sample(x=c(TRUE,FALSE),size = nrow(ratings_movies),replace = TRUE,prob = c(.8,.2))
rec_data_train<-ratings_movies[which_train,]
rec_data_test<-ratings_movies[!which_train,]

#using the training dataset to create recommender model
rec <- Recommender(rec_data_train, method = "IBCF")

#predict recommended Movie for  users on test dataset
rec_pred <- predict(rec, rec_data_test, n=10)
rec_pred
pre<-sapply(rec_pred@items,
            function(x){
              colnames(ratings_movies)[x]
            })

```
